---
title: "bikes"
author: Matthew Parin
output: html_document
---

Source: http://pedrosan.github.io/DataScience/Bike_Sharing/

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Install packages that are needed:

```{r, eval=FALSE}
# install.packages("lubridate")
# install.packages("data.table")
# install.packages("cowplot")
install.packages("plyr")
```

Load the package functions into your environment:

```{r, eval=TRUE}
library(data.table)
library(lubridate)
library(cowplot)
library(plyr)
```

## Read the data into the environment

```{r, eval=FALSE}
stations <- as.data.frame(fread("~/DivvyChicago/data/Divvy_Stations_2017_Q1Q2.csv",
                                sep = ",", header = TRUE, stringsAsFactors = FALSE))
```

```{r, eval=FALSE}
rides <- as.data.frame(fread("~/DivvyChicago/data/Divvy_Trips_2016.csv",
                             sep = ",", header = TRUE, stringsAsFactors = FALSE))
```

## Preparing the data for analysis

Preparing a smaller set of data

```{r}
data <- rides[, c("bikeid", "from_station_id", "to_station_id", "starttime", "stoptime", "tripduration")]
# data$starttime <- mdy_hms(data$starttime, tz = "US/Central")
# data$stoptime <- mdy_hms(data$stoptime, tz = "US/Central")
```

Sort the data by bike and start time and reorder the row names

```{r}
initial.sort <- order(data$bikeid, data$starttime)
data <- data[initial.sort, ]
row.names(data) <- NULL
```

What's the trip duration summary

```{r}
summary(as.numeric(data$tripduration))
```

Let's inspect the data a little more.

```{r}
# checking for NA
sum(is.na(data))

# checking number of "levels" 
length(unique(data$bikeid))
length(unique(data$from_station_id))
length(unique(data$to_station_id))

# create arrays of unique values for bike, station1, station2 IDs
bikes <- unique(data$bikeid)
from_station_id <- unique(data$from_station_id)
to_station_id <- unique(data$to_station_id)

# check cross-matching of station lists
sum(! (data$from_station_id %in% data$to_station_id) )  # 1 false ?
sum(! (data$to_station_id %in% data$from_station_id) )  # all true
```

What's the minimum number of mismatches -- i.e. folks moving bikes around

```{r}
# finding mis-matching stations over the full data set
bike.stats <- ddply(data, .(bikeid), summarize, 
                    Ntrips = length(starttime),
                    mismatches = sum(from_station_id[-1] != to_station_id[1:(length(starttime)-1)]), 
                    fraction = mismatches/Ntrips)

number.mismatches <- sum(bike.stats$mismatches) 

# Minimum-Original-Total = Existing data + Number of mismatches
minimum.missing.percent <- (number.mismatches / (nrow(data) + number.mismatches))*100
print(minimum.missing.percent)
```

It looks like ~18% of the bikes are moved around the city to manage supply and demand.

Can we figure out how long bikes stay at their stations?

```{r}
# main <- data[-nrow(data), ]
# main$tdiff <- difftime(data$starttime[-1], data$stoptime[-nrow(data)])
# main$flag.combined <- (data$from_station_id[-1] == data$to_station_id[-nrow(data)]) & 
#                       (data$bikeid[-1] == data$bikeid[-nrow(data)])
# main$flag.station <- c(main$from_station_id[-1] == main$to_station_id[-nrow(main)], NA)
# 
# good.tdiff <- main$tdiff[main$flag.combined & as.numeric(main$tdiff) > 0]
# length(good.tdiff)
```





