---
title: "Fleet Performance"
author: "Matthew Parin"
date: "8/30/2017"
output: html_document
---

Source: https://docs.google.com/spreadsheets/d/1phtN9ptIgmMAO0tNcf6LyJkh39UNNAmZXTQ1oqpCQo0/edit#gid=0

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
current.year <- as.integer(format(Sys.Date(), "%Y"))
# mycolors <-  c("#CC0000", "#666666", "#009E73", "#CCCCCC", "#F0E442", "#0072B2", "#CC79A7")
coord <- as.data.frame(geocode("Chicago, Illinois", output = "latlon"))
```

Install packages that are needed:

```{r, eval=FALSE}
# install.packages("lubridate")
# install.packages("data.table")
# install.packages("cowplot")
# install.packages("data.table")
# install.packages("plyr")
# install.packages("dplyr")
# install.packages("scales")
# install.packages("tidyr")
# install.packages("DT")
# install.packages("ggthemes")
# install.packages("ggplot2")
# install.packages("ggmap")
```

Load the package functions into your environment:

```{r, eval=TRUE}
library(data.table)
library(lubridate)
library(cowplot)
library(plyr)
library(dplyr)
library(scales)
library(tidyr)
library(DT)
library(ggthemes)
library(ggplot2)
library(ggmap)
```

## Read the data into the environment

```{r, eval=FALSE}
stations <- as.data.frame(fread("~/DivvyChicago/data/Divvy_Stations_2017_Q1Q2.csv",
                                sep = ",", header = TRUE, stringsAsFactors = FALSE))
```

```{r, eval=FALSE}
rides <- as.data.frame(fread("~/DivvyChicago/data/Divvy_Trips_2016.csv",
                             sep = ",", header = TRUE, stringsAsFactors = FALSE))
```

## Clean the Station data

```{r, eval=FALSE}
rownames(stations) <- stations$name
stations$name      <- NULL
```

## Clean the Ride data

Create a function for parsing date and time
```{r, eval=TRUE}
parse.datettime <- function (x) {
  out    <- suppressWarnings(mdy_hms(x))
  i      <- is.na(out)
  out[i] <- mdy_hm(x[i])
  return(out)
}
```

Use transform() to add columns for bike pickup and dropoff
```{r, eval=FALSE}
rides <- transform(rides, usertype = factor(usertype))
rides <- transform(rides, gender = factor(gender))
rides <- transform(rides, birthyear = as.numeric(birthyear))
rides <- transform(rides, current.age = as.numeric(current.year - birthyear))
rides <- transform(rides,
                   from_station_name = factor(from_station_name,
				                              rownames(stations)),
                   to_station_name   = factor(to_station_name,
				                              rownames(stations)))

rides <- transform(rides, starttime = parse.datettime(starttime))
rides <- transform(rides, start.hour = as.numeric(format(starttime, "%H")))
rides <- transform(rides, start.minute = as.numeric(format(starttime, "%M")))
rides <- transform(rides, start.second = as.numeric(format(starttime, "%S")))
rides <- transform(rides, start.date = as.numeric(format(starttime, "%d")))
rides <- transform(rides, start.week = as.numeric(format(starttime,"%W")))
rides <- transform(rides, start.week = factor(start.week))
rides <- transform(rides,
           start.day = factor(weekdays(as.Date(starttime)),
                              c("Monday","Tuesday","Wednesday","Thursday",
                                "Friday","Saturday","Sunday")))
rides <- transform(rides,
           start.month = factor(months(as.Date(starttime)),
                              c("January","February","March","April",
                                "May", "June", "July", "August",
                                "September","October","November",
                                "December")))
rides <- transform(rides, day.year = as.numeric(strftime(starttime, format = "%j")))

rides <- transform(rides, stoptime = parse.datettime(stoptime))
rides <- transform(rides, stop.hour = as.numeric(format(stoptime, "%H")))
rides <- transform(rides, stop.minute = as.numeric(format(stoptime, "%M")))
rides <- transform(rides, stop.second = as.numeric(format(stoptime, "%S")))
rides <- transform(rides, stop.date = as.numeric(format(stoptime, "%d")))
rides <- transform(rides, stop.week = as.numeric(format(stoptime,"%W")))
rides <- transform(rides, stop.week = factor(stop.week))
rides <- transform(rides,
           stop.day = factor(weekdays(as.Date(stoptime)),
                              c("Monday","Tuesday","Wednesday","Thursday",
                                "Friday","Saturday","Sunday")))
rides <- transform(rides,
           stop.month = factor(months(as.Date(stoptime)),
                              c("January","February","March","April",
                                "May", "June", "July", "August",
                                "September","October","November",
                                "December")))
rides <- transform(rides, day.year.stop = as.numeric(strftime(stoptime, format = "%j")))
```

## Prepare KPI data on fleet

# Fleet Size

Maximum Bikes on Street
```{r}
number.bikes <- length(unique(rides$bikeid))
print(number.bikes)
```

Number of Stations
```{r}
number.stations <- length(unique(stations$id))
print(number.stations)
```

Number of Docks
```{r}
number.docks <- sum(as.integer(stations$dpcapacity))
print(number.docks)
```

# Customer Size 

Annual Members
```{r}
subscribers <- subset(rides, rides$usertype == "Subscriber")
dependents <- subset(rides, rides$usertype == "Dependent")
```

24-hour Members
```{r}
customers <- subset(rides, rides$usertype == "Customer")
```

# Trip Metrics

```{r}
by_user_date <- rides %>%
  group_by(usertype, day.year) %>%
  dplyr::summarise(Total = n())

average.by_user_date <- by_user_date %>%
  group_by(usertype) %>%
  dplyr::summarize(mean = mean(Total))

print(average.by_user_date)
```


Total Trips
```{r}
number.trips <- length(rides$trip_id)
```

Peak Month Average Trips Per Bike Per Day
```{r}
by_month_date_bike <- rides %>%
  group_by(start.month, day.year, bikeid) %>%
  dplyr::summarize(Total = n())

peak_average.by_month_date_bike <- by_month_date_bike %>%
  group_by(start.month) %>%
  dplyr::summarize(mean = mean(Total))

print(max((peak_average.by_month_date_bike$mean)))
```

Average Trips Per Bike Per Day 
```{r}
by_day_bike <- rides %>%
  group_by(start.month, day.year, bikeid) %>%
  dplyr::summarize(Total = n())

average.by_date_bike <- by_day_bike %>%
  group_by(day.year) %>%
  dplyr::summarize(mean = mean(Total))

print(mean(average.by_date_bike$mean))
```

Average Trip Length (Miles)
```{r}

```

Average Trip Duration
```{r}
seconds_to_period(mean(as.numeric(rides$tripduration)))
```

Total Number of Trips by Hour
```{r}
by_hour <- rides %>%
  group_by(start.hour) %>%
  dplyr::summarize(Total = n())

datatable(by_hour)
```

Plot Total Number of Divvy Trips by Hour
```{r}
ggplot(by_hour, aes(start.hour, Total)) +
  geom_bar(stat = "identity", fill = "darkgreen") +
  ggtitle("Total 2016 Divvy Trips Every Hour") +
  theme(legend.position = "none") +
  scale_y_continuous(labels = comma)
```

Total Number of Divvy Trips by Hour (monthly)
```{r}
by_month_hour <- rides %>%
  group_by(start.month, start.hour) %>%
  dplyr::summarize(Total = n())
```

Plot Total Number of Divvy Trips by Hour (monthly)
```{r}
ggplot(by_month_hour, aes(start.hour, Total, fill = start.month)) +
  geom_bar(stat = "identity") +
  ggtitle("Total 2016 Divvy Trips by Hour and Month") +
  scale_y_continuous(labels = comma) # + you need 12 colors to make the distinction
  # scale_fill_manual(values = mycolors)
```

Define start and stop locations
```{r}
# station.locations <- data.frame(station_id = stations$id, lon=stations$longitude, lat=stations$latitude)
```






