---
title: "Navy Pier Rides"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Visualizing the usage patterns of Navy Pier Divvy users

```{r, eval=FALSE}
summary(navy_pier.rides$to_station_name)
```

What could explain this pattern? One hypothesis is that geography can
explain this trend---to investigate, let's visualize these data.

First, create a variable storing counts of the number of trips by
arrival station.

```{r, eval=FALSE}
from.navy_pier.counts <- as.vector(table(navy_pier.rides$to_station_name))
head(from.navy_pier.counts)
```

Many of the stations in the city are accessed from Navy Pier.

```{r, eval=FALSE}
summary(factor(from.navy_pier.counts == 0))
```

Next, let's create another variable with the (Euclidean) distance
between the Navy Pier station and all other stations across the
city.

```{r, eval=FALSE}
navy_pier <- "Streeter Dr & Grand Ave"
navy_pier.dists <-
  sqrt((stations[navy_pier,"latitude"] - stations$latitude)^2 +
       (stations[navy_pier,"longitude"] - stations$longitude)^2)
head(navy_pier.dists)
```

To compare the counts and distance, let's combine these two variables
into a single data frame. Once we have a data frame, we can plot the
relationship between these two variables using ggplot.

```{r, eval=FALSE}
from.navy_pier <- data.frame(dist  = navy_pier.dists,
                            count = from.navy_pier.counts)
rownames(from.navy_pier) <- rownames(stations)
head(from.navy_pier)
```

Note that we are able to combine these data into one data frame only
because we were careful earlier to create factors with levels given in
the same order as the stations in the "stations" data frame.

Let's create a scatterplot comparing the two variables.

```{r, eval=FALSE}
library(ggplot2)
ggplot(data    = from.navy_pier,
       mapping = aes(x = dist,y = count)) +
  geom_point()
```

This plot isn't particularly informative because a few trip counts are
much larger than the others, and a few of the distances are much
smaller. This suggests that a log-scale will work better.

```{r, eval=FALSE}
ggplot(data    = from.navy_pier,
       mapping = aes(x = dist,y = count)) +
  geom_point() +
  scale_x_log10() +
  scale_y_log10()
```

That's better, but we get a couple warnings because we are taking the
log of zero. Let's fix this by removing all points at zero.

```{r, eval=FALSE}
ggplot(data    = subset(from.navy_pier, count > 0 & dist > 0),
       mapping = aes(x = dist, y = count)) +
  geom_point(shape = 4, color = "darkblue") +
  scale_x_log10() +
  scale_y_log10(breaks = c(1,10,100,1000))
```

That's a big improvement. One final adjustment: let's add more
informative axis labels. And then we can save the plot for later.

```{r, eval=FALSE}
plot3 <- ggplot(data    = subset(from.navy_pier, count > 0 & dist > 0),
                mapping = aes(x = dist,y = count)) +
  geom_point(shape = 4,color = "darkblue") +
  scale_x_log10() +
  scale_y_log10(breaks = c(1,10,100,1000)) +
  labs(x = "Distance from Navy Pier",
       y = "Number of Arrivals")
print(plot3)
```

This plot clearly shows that the number of trips to each station drops
off linearly with distance from the station (on the log-scale). It
would be interesting to see whether a similar trend occurs city-wide.

### Analyzing the linear relationship between two variables

Let's analyze the linear relationship between trip counts and geographic
distance.

```{r, eval=FALSE}
fit <- lm(data    = subset(from.navy_pier, count > 0 & dist > 0),
          formula = log10(count) ~ log10(dist))
```

As before, we can get a detailed summary of the linear regression
using the "summary" function:

```{r, eval=FALSE}
class(fit)
summary(fit)
```

Among other things, the summary tells us that about 50% of the
variation in trip arrival counts is explained by geographic distance.

We can also add the fitted regression line to the scatterplot we
created earlier:

```{r, eval=FALSE}
b <- coef(fit)
plot3 + geom_abline(intercept = b["(Intercept)"],slope = b["log10(dist)"],
                    color = "red",linetype = "dotted")
```
