---
title: "Divvy Revenue (2016)"
author: "Matthew Parin"
date: "September 5, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Install packages that are needed:
```{r, eval=FALSE}
# install.packages("lubridate")
# install.packages("data.table")
# install.packages("plyr")
# install.packages("dplyr")
# install.packages("tidyr")
# install.packages("DT")
# install.packages("ggplot2")
```

Load the package functions into your environment:
```{r, eval=TRUE}
library(lubridate)
library(data.table)
library(plyr)
library(dplyr)
library(scales)
library(tidyr)
library(DT)
library(ggplot2)
```

## Read the Ride data into the environment

```{r, eval=FALSE}
rides <- as.data.frame(fread("~/DivvyChicago/data/Divvy_Trips_2016.csv",
                             sep = ",", header = TRUE, stringsAsFactors = FALSE))
```

# Clean the Rider data

```{r, eval=FALSE}
rides$tripduration <- as.numeric(rides$tripduration)
```


## Context of the Divvy Revenue Model

If you are a General Manager of a product with multiple revenue streams, you have to think through different evaluation models meant to define progress toward your business objectives.

For the purposes of this model, I want to simply define how much of the Chicago Divvy revenue comes from rides of various types. I draw all my pricing data from https://www.divvybikes.com/pricing/ and use the trip duration of the 2016 Divvy rides data to inform the analysis.

To date, my model includes information on:
- Gross revenue
- Gross revenue, by user type
- Gross revenue, by age segment (TODO)
- Average revenue, by user type
- Average revenue, by age segment (TODO)
- Daily average revenue of subscribers

Revenue models inspired by https://www.marketingprofs.com/articles/2017/32037/revenue-is-not-a-kpi-but-these-six-measures-are

# Create a new data frame for analysis
```{r, eval=FALSE}
rides.df <- rides[, c("bikeid", "tripduration", "usertype", "gender", "birthyear")]
```

# Calculate the age of the user
```{r, eval=FALSE}
data.year <- 2016
rides.df <- transform(rides.df, birthyear = as.numeric(birthyear))
rides.df <- transform(rides.df, current.age = as.numeric(data.year - birthyear))
```

# Segment the data into similar priced groups
I want to remove the "Dependent" segment, as it appears to be a single user
```{r, eval=FALSE}
rides.subscribers <- subset(rides.df, rides.df$usertype == "Subscriber")
rides.customers <- subset(rides.df, rides.df$usertype == "Customer")
rides.dependent <- subset(rides.df, rides.df$usertype == "Dependent")
rm(rides.dependent)
```

# Trip revenue function for Subscribers and Customers (Test-driven data analysis)
```{r, eval=FALSE}
# // Create mock data
# rides.duration.test <- as.data.frame(matrix(c(5500, 3700, 1900, 1500),
#                                 nrow = 4,
#                                 ncol = 1))

# // Iterate on the function
# ride.revenue.subscriber <- function(x) {
#   if (x > 5400) {
#     result <- "High" #1.5 + 4.5 + (((x-5400) / 1800) * 6)
#   }
#   else if (3600 < x && x <= 5400) {
#     result <- "Mid" #1.5 + 4.5
#   }
#   else if (1800 < x && x <= 3600) {
#     result <- "Low" #1.5
#   }
#   else {
#     result <- "No" #0
#   }
#   return(result)
# }

# // Test the results
# rides.duration.test$cost <- round(apply(rides.duration.test[1:1], 1, ride.revenue.subscriber), digits = 2)
```

# Trip revenue function for Subscribers and Customers
```{r, eval=FALSE}
ride.revenue.subscriber <- function(x) {
  if (x > 5400) {
    result <- 1.5 + 4.5 + (((x-5400) / 1800) * 6)
  }
  else if (3600 < x && x <= 5400) {
    result <- 1.5 + 4.5
  }
  else if (1800 < x && x <= 3600) {
    result <- 1.5
  }
  else {
    result <- 0
  }
  return(result)
}

ride.revenue.customer <- function(x) {
  if (x > 5400) {
    result <- 2 + 6 + (((x-5400) / 1800) * 8)
  }
  else if (3600 < x && x <= 5400) {
    result <- 2 + 6
  }
  else if (1800 < x && x <= 3600) {
    result <- 2
  }
  else {
    result <- 0
  }
  return(result)
}
```

# Find the cost of a ride
```{r, eval=FALSE}
rides.subscribers$cost <- round(apply(rides.subscribers[2:2], 1,
                                      ride.revenue.subscriber), digits = 2)
rides.customers$cost <- round(apply(rides.customers[2:2], 1,
                                    ride.revenue.customer), digits = 2)
```

# Calculate trip revenue
```{r, eval=TRUE}
subscriber.trip.revenue <- sum(rides.subscribers$cost)
customer.trip.revenue <- sum(rides.customers$cost)
total.trip.revenue <- subscriber.trip.revenue + customer.trip.revenue
print(total.trip.revenue)
```

# Define assumptions for revenue analysis
```{r, eval=FALSE}
# Source: http://www.chicagotribune.com/news/local/breaking/ct-divvy-rides-record-met-20161229-story.html
# Assumption: International Bike-share database is accurate

# Membership
subscribers <- 33586
customers <- 354008

# Membership not covered in data
student <- 2074
equity <- 1965

# Fees
cost.subscription <- 99
cost.customer <- 9.95

# Ridership
average.rides_customer <- length(rides.customers$usertype) / customers
average.rides_subscriber <- length(rides.subscribers$usertype) / subscribers
```

# Calculate fees revenue
```{r, eval=TRUE}
subscriber.fees.revenue <- subscribers*cost.subscription
customer.fees.revenue <- (length(rides.customers$usertype)/average.rides_customer)*cost.customer
total.fees.revenue <- subscriber.fees.revenue + customer.fees.revenue
print(total.fees.revenue)
```

# Calculate total revenue
```{r, eval=TRUE}
total.gross.revenue <- total.trip.revenue + total.fees.revenue
print(total.gross.revenue)
```

# Combine separate segments into a single data frame
```{r, eval=FALSE}
rides.df.revenue <- rbind(rides.subscribers, rides.customers)
```

Calculate financial indicators
```{r, eval=TRUE}
# Average revenue by user type
average.revenue.subscriber <- (subscriber.fees.revenue + subscriber.trip.revenue) / subscribers
average.revenue.customer <- (customer.fees.revenue + customer.trip.revenue) / customers

# Daily average revenue by user type
daily.average.revenue.subscriber <- average.revenue.subscriber / 365
daily.average.revenue.customer <- average.revenue.customer

# Average ride revenue for individual rides by user type
average.revenue.ride.subscriber_per.trip <- sum(rides.subscribers$cost) /
  length(rides.subscribers$cost)
average.revenue.ride.customer_per.trip <- sum(rides.customers$cost) / 
  length(rides.customers$cost)
average.revenue.ride.total_per.trip <- sum(rides.df.revenue$cost) /
  length(rides.df.revenue$cost)

# Average ride revenue for combined rides by user type
average.revenue.ride.subscriber_combined <- sum(rides.subscribers$cost) /
  subscribers
average.revenue.ride.customer_combined <- sum(rides.customers$cost) / 
  customers
average.revenue.ride.total_combined <- sum(rides.df.revenue$cost) /
  (subscribers + customers)
```

# Plot the average revenue of subscribers, by age
```{r, eval=TRUE}
library(ggplot2)
ride.subscibers_plot <- rides.subscribers %>%
  group_by(gender, current.age) %>%
  dplyr::summarize(mean = mean(cost))

ride.subscibers_plot <- subset(ride.subscibers_plot, ride.subscibers_plot$gender != "")

ride.subscibers_plot$mean_z = (ride.subscibers_plot$mean -
  mean(ride.subscibers_plot$mean)) /
  sd(ride.subscibers_plot$mean)

ride.subscibers_plot$type <- ifelse(ride.subscibers_plot$mean_z < 0, "below", "above")
ride.subscibers_plot <- ride.subscibers_plot[order(ride.subscibers_plot$current.age), ]
ride.subscibers_plot$current.age <- factor(ride.subscibers_plot$current.age)

ggplot(ride.subscibers_plot, aes(x=current.age, y=mean_z, label=mean_z)) + 
  geom_bar(stat = 'identity', aes(fill=gender), width=.5)  +
  labs(subtitle="Subdivided by Gender", 
       title= "Relative Value of Divvy Users by Age") + 
  ylim(-2.5, 2.5) +
  coord_flip()
```

# Segment Subscribers by age
```{r, eval=FALSE}
age.labels <- c("Child", "College", "Young Professional", "30s", "40s", "50s", "60s", "70s", "80s", "90s", "100+")
rides.subscribers$age.segments <- cut(
  rides.subscribers$current.age,
  breaks = c(0, 17, 22, 29, 39, 49, 59, 69, 79, 89, 99, 149),
  right = FALSE,
  labels = age.labels
)
```

# Plot the age segment data
```{r, eval=TRUE}
library(ggplot2)
# Create the data frame for analysis
ride.subscibers_age.plot <- rides.subscribers %>%
  group_by(gender, age.segments) %>%
  dplyr::summarize(mean = mean(cost))

# Remove the files that do not include gender
ride.subscibers_age.plot <- subset(ride.subscibers_age.plot, ride.subscibers_age.plot$gender != "")

# Add a column with a Z score
ride.subscibers_age.plot$mean_z = (ride.subscibers_age.plot$mean -
  mean(ride.subscibers_age.plot$mean)) /
  sd(ride.subscibers_age.plot$mean)

# Prepare the data for plotting
ride.subscibers_age.plot$type <- ifelse(ride.subscibers_age.plot$mean_z < 0, "below", "above")
ride.subscibers_age.plot <- ride.subscibers_age.plot[order(ride.subscibers_age.plot$mean_z), ]
ride.subscibers_age.plot$age.segments <- factor(ride.subscibers_age.plot$age.segments)

# Plot the data
ggplot(ride.subscibers_age.plot, aes(x=age.segments, y=mean_z, label=mean_z)) + 
  geom_bar(stat = 'identity', aes(fill=gender), width=.5)  +
  labs(subtitle="Subdivided by Gender", 
       title= "Relative Value of Divvy Users by Age") + 
  ylim(-2.5, 2.5) +
  coord_flip()
```

## Analysis

The revenue model of Divvy is dominated by customer fees. Nearly 45% of Divvy's annual revenue is composed of 24-hour subscription fees paid by those using Divvy for a short time, based on assumptions.

Moreover, fees from customers and subscribers dominate nearly 80% of Divvy's 9.65M USD gross revenue. This is a likely outcome given almost 92% of rides do not incur additional time-based fees, for rides which extend beyond 30 minutes.

In terms of straight revenue, the revenue value of each subscriber is nearly 8x each short-term customer. Subscribers bring in 109.78 USD of revenue per year whereas customers bring in 13.89 per day.

On an equal basis, customer bring in 13.89 USD per day and subscribers bring in 0.30 USD per day. This means the question of whether or not to close down the system at any given time throughout the year (i.e. during the winter), turns into an economic question of whether delivering bikes at a price of 14.19 USD per day is above the average total cost (ATC) of delivering bikes in a specific period. Divvy does not shut down during the winter, so it follows the marginal cost (MC) of delivering bikes every day of the year is above ATC.