---
title: "Divvy Revenue (2016)"
author: "Matthew Parin"
date: "September 5, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
current.year <- as.integer(format(Sys.Date(), "%Y"))
# coord <- as.data.frame(geocode("Chicago, Illinois", output = "latlon"))
```

Install packages that are needed:
```{r, eval=FALSE}
# install.packages("lubridate")
# install.packages("data.table")
# install.packages("cowplot")
# install.packages("data.table")
# install.packages("plyr")
# install.packages("dplyr")
# install.packages("scales")
# install.packages("tidyr")
# install.packages("DT")
# install.packages("ggthemes")
# install.packages("ggplot2")
# install.packages("ggmap")
```

Load the package functions into your environment:
```{r, eval=TRUE}
library(data.table)
library(lubridate)
library(cowplot)
library(plyr)
library(dplyr)
library(scales)
library(tidyr)
library(DT)
library(ggthemes)
library(ggplot2)
library(ggmap)
```

## Read the Ride data into the environment

```{r, eval=FALSE}
rides <- as.data.frame(fread("~/DivvyChicago/data/Divvy_Trips_2016.csv",
                             sep = ",", header = TRUE, stringsAsFactors = FALSE))
```

# Clean the Rider data

```{r, eval=FALSE}
rides$tripduration <- as.numeric(rides$tripduration)
```


## Context of the Divvy Revenue Model

If you are a General Manager of a product with multiple revenue streams, you have to think through how best to adjust prices to achieve your business objectives.

For the purposes of this model, I want to simply define how much of the Chicago Divvy revenue comes from rides of various types. I draw all my pricing data from https://www.divvybikes.com/pricing/ and use the trip duration of the 2016 Divvy rides data to inform the analysis.

Segment the data into similar priced groups
```{r, eval=FALSE}
rides.subscribers <- subset(rides, rides$usertype == "Subscriber")
rides.customers <- subset(rides, rides$usertype == "Customer")
rides.dependent <- subset(rides, rides$usertype == "Dependent")
```

Pricing specifics for Subscribers and Customers
```{r, eval=FALSE}
# rides.duration.sample <- as.data.frame(matrix(c(5500, 3700, 1900, 1500),
#                                 nrow = 4,
#                                 ncol = 1))

# ride.revenue.subscriber <- function(x) {
#   if (x > 5400) {
#     result <- "High" #1.5 + 4.5 + (((x-5400) / 1800) * 6)
#   }
#   else if (3600 < x && x <= 5400) {
#     result <- "Mid" #1.5 + 4.5
#   }
#   else if (1800 < x && x <= 3600) {
#     result <- "Low" #1.5
#   }
#   else {
#     result <- "No" #0
#   }
#   return(result)
# }

ride.revenue.subscriber <- function(x) {
  if (x > 5400) {
    result <- 1.5 + 4.5 + (((x-5400) / 1800) * 6)
  }
  else if (3600 < x && x <= 5400) {
    result <- 1.5 + 4.5
  }
  else if (1800 < x && x <= 3600) {
    result <- 1.5
  }
  else {
    result <- 0
  }
  return(result)
}

ride.revenue.customer <- function(x) {
  if (x > 5400) {
    result <- 2 + 6 + (((x-5400) / 1800) * 8)
  }
  else if (3600 < x && x <= 5400) {
    result <- 2 + 6
  }
  else if (1800 < x && x <= 3600) {
    result <- 2
  }
  else {
    result <- 0
  }
  return(result)
}
```

Find the cost of a ride
```{r}
# rides.duration.sample$cost <- round(apply(rides.duration.sample[1:1], 1, ride.revenue.subscriber), digits = 2)

rides.dependent$cost <- round(apply(rides.dependent[5:5], 1, ride.revenue.subscriber), digits = 2)
rides.subscribers$cost <- round(apply(rides.subscribers[5:5], 1, ride.revenue.subscriber), digits = 2)
rides.customers$cost <- round(apply(rides.customers[5:5], 1, ride.revenue.customer), digits = 2)
```

Calculate trip revenue
```{r}
subscriber.trip.revenue <- sum(rides.subscribers$cost)
dependent.trip.revenue <- sum(rides.dependent$cost)
customer.trip.revenue <- sum(rides.customers$cost)
total.trip.revenue <- subscriber.trip.revenue + dependent.trip.revenue + customer.trip.revenue
print(total.trip.revenue)
```

Calculate fees revenue
```{r}
subscriber.fees.revenue <- 33600*99
customer.fees.revenue <- length(rides.customers$usertype)*9.95
total.fees.revenue <- subscriber.fees.revenue + customer.fees.revenue
print(total.fees.revenue)
```

Calculate total revenue
```{r}
total.gross.revenue <- total.trip.revenue + total.fees.revenue
print(total.gross.revenue)
```

