---
title: "Analysis Setup"
author: "Matthew Parin"
output: html_document
---

## Setup instructions

```{r, echo=FALSE, results="hide", eval=TRUE, message=FALSE}
library(data.table)
library(lubridate)
library(ggplot2)
library(cowplot)
```

To work through the lessons on your computer, please follow these
setup instructions first.

1. Install R from the [CRAN website](http://cran.r-project.org) and, optionally,
[RStudio Desktop](https://www.rstudio.com/products/rstudio/download).
*Note:* If you do not use RStudio Desktop, you will need to have
additional software installed for X Windows graphics (e.g.,
[XQuartz](http://xquartz.org) on Mac). Without this, you will not be
able to create plots.

2. Download or clone the [git repository](https://github.com/matthewparin/DivvyChicago).
Click [here](https://github.com/matthewparin/DivvyChicago/archive/master.zip)
to download a Zip file containing all the files.

3. Download the Divvy data files into the "data" directory, and
prepare the data files for analysis. **Easier approach:** download
the data directly from the Divvy website. I have provided a bash script
`scripts/get_2016_Divvy_Data.R` and `scripts/get_2017_Divvy_Data.R`
to do this automatically. This script reproduces the data processing
steps starting with the [original source](https://www.divvybikes.com/system-data).

4. After completing Step 4, check that you have all the necessary data
files in the "data" directory. For example, on my MacBook Air, I have:

```{r}
list.files("../data")
```

5. Launch R or RStudio Desktop.

6. Make sure your working directory in R or RStudio is set to the
"files" directory of the git repository. For example, on my MacBook
Air, the current working directory in R is

```{r}
getwd()
```

If you are not in the files directory, use the "setwd" function or
the RStudio menus to change the working directory.

Once you have completed these steps, you are ready to move on to the
lessons.

## Session information

The "sessionInfo" function shows the version of R I am using, and the
packages that I used to generate the webpages from the R Markdown
documents.

```{r session-info}
```